Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-06-02 source/sec_Nesterov.tex
minor edits
updated nesterov statement
updated sgd and nesterov

\section{Linear Speedup Analysis of Nesterov Accelerated FedAvg}

A natural extension of the FedAvg algorithm is to use momentum-based
local updates instead of local SGD updates. To our knowledge, the
only convergence analyses of FedAvg with momentum-based stochastic
updates focus on the non-convex smooth case \cite{yu2019linear,huo2020faster,li2018federated}.
In this section, we complete the picture with $\mathcal{O}(1/NT)$
and $\mathcal{O}(1/\sqrt{NT})$ convergence results for Nesterov-accelerated
FedAvg for convex objectives. As we know from stochastic optimization,
Nesterov updates may fail to accelerate over SGD, so in general we
cannot hope to obtain acceleration results over FedAvg. However, in
the following section we specialize to overparameterized problems
where we demonstrate that a particular FedAvg variant with momentum
updates is able to accelerate over the original FedAvg algorithm.

@@

We first show that the Nesterov accelerated FedAvg has $\mathcal{O}(1/NT)$
convergence rate for $\mu$-strongly convex and $L$-smooth objectives.
The Nesterov Accelerated FedAvg algorithm follows the updates
\begin{align*}
\mathbf{v}_{t+1}^{k} & =\mathbf{w}_{t}^{k}-\alpha_{t}g_{t,k}\\
\mathbf{w}_{t+1}^{k} & =\begin{cases}
\mathbf{v}_{t+1}^{k}+\beta_{t}(\mathbf{v}_{t+1}^{k}-\mathbf{v}_{t}^{k}) & \text{if }t+1\notin\mathcal{I}_{E}\\
\sum_{k=1}^{N}p_{k}\left[\mathbf{v}_{t+1}^{k}+\beta_{t}(\mathbf{v}_{t+1}^{k}-\mathbf{v}_{t}^{k})\right] & \text{if }t+1\in\mathcal{I}_{E}
\end{cases}
\end{align*}
where $g_{t,k}:=\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})$ is
the stochastic gradient sampled on the $k$-th device at time $t$.  
\begin{theorem}
	\label{thm:nesterov_scvx}Let $\overline{\mathbf{v}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{v}_{T}^{k}$
	be the average of local parameters of Nesterov accelerated FedAvg
	with learning rates $\alpha_{t}=\frac{6}{\mu}\frac{1}{t+\gamma}$
	and $\beta_{t-1}=\frac{3}{14(t+\gamma)(1-\frac{6}{t+\gamma})\max\{\mu,1\}}$,
	$\gamma=\max\{32\kappa,E\}$ and $\kappa=\frac{L}{\mu}$. Then with
	full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}=\mathcal{O}(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}})
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}\leq\mathcal{O}(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}})
	\end{align*}
\end{theorem}
%
\begin{remark}
To our knowledge, this is the first convergence result for Nesterov
accelerated FedAvg in the strongly convex and smooth setting. The
same discussion about linear speedup of FedAvg applies to the Nesterov
accelerated variant. In particular, to achieve $\mathcal{O}(1/NT)$
linear speedup, $T$ iterations of the algorithm requires only $\mathcal{O}(\sqrt{NT})$
communication rounds with full participation. 
\end{remark}

@@

We now provide linear speedup analysis of Nesterov Accelerated FedAvg
with convex and smooth objectives and show that the optimality gap
is $\mathcal{O}(1/\sqrt{NT})$ where $N$ is the number of participating
devices. This result complements the strongly convex case in the previous
part, as well as the non-convex smooth setting in \cite{huo2020faster,yu2019linear,li2018federated},
where a similar $\mathcal{O}(1/\sqrt{NT})$ rate is given in terms
of averaged gradient norm. 
\begin{theorem}
	\label{thm:Nesterov_cvx}Set learning rates $\alpha_{t}=\beta_{t}=\mathcal{O}(\sqrt{\frac{N}{T}})$
	for Accelerated FedAvg with convex and $L$-smooth objectives. With
	full device participation, 
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{w}}_{t})-F^{\ast} & =\mathcal{O}(\sqrt{\frac{\nu_{\max}^{2}\sigma^{2}}{NT}+\frac{E^{2}LG^{2}}{T^{4/3}}})
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{w}}_{t})-F^{\ast} & =\mathcal{O}(\sqrt{\frac{\nu_{\max}^{2}\sigma^{2}}{NT}+\frac{E^{2}G^{2}}{KT}+\frac{E^{2}LG^{2}}{T^{4/3}}})
	\end{align*}
\end{theorem}
%
\begin{remark}
	It is possible to extend the analyses in this section to accelerated
	FedAvg algorithms with other momentum-based updates. However, in the
	stochastic optimization setting, none of these methods will be able
	to achieve a better rate than the original FedAvg with SGD in general.
	For this reason, we will instead turn to the overparameterized setting
	\cite{ma2017power,liu2018accelerating,canziani2016analysis} in the
	next section where we show that FedAvg enjoys exponential convergence
	and it is possible to improve its convergence rate with momentum based
	local updates. 
\end{remark}

>>>>

\section{Linear Speedup Analysis of Nesterov Accelerated FedAvg}
\label{sec:Nesterov}

A natural extension of the FedAvg algorithm is to use momentum-based
local updates instead of local SGD updates. To our knowledge, the
only convergence analyses of FedAvg with momentum-based stochastic
updates focus on the non-convex smooth case \cite{huo2020faster,yu2019linear,li2018federated}.
In this section, we complete the picture with $\mathcal{O}(1/NT)$
and $\mathcal{O}(1/\sqrt{NT})$ convergence results for Nesterov-accelerated
FedAvg for convex objectives that match the original FedAvg results.
As we know from stochastic optimization, Nesterov and other momumtum
updates may fail to accelerate over SGD \cite{liu2018accelerating,kidambi2018insufficiency,liu2018toward,yuan2016influence}.
Therefore in Section~\ref{sec:overparameterized} we will specialize to overparameterized
problems where we demonstrate that a particular FedAvg variant with
momentum updates is able to accelerate over the original FedAvg algorithm.

@@

We first show that the Nesterov accelerated FedAvg has $\mathcal{O}(1/NT)$
convergence rate for $\mu$-strongly convex and $L$-smooth objectives.
The Nesterov Accelerated FedAvg algorithm follows the updates
\begin{align*}
\mathbf{v}_{t+1}^{k} & =\mathbf{w}_{t}^{k}-\alpha_{t}g_{t,k}\\
\mathbf{w}_{t+1}^{k} & =\begin{cases}
\mathbf{v}_{t+1}^{k}+\beta_{t}(\mathbf{v}_{t+1}^{k}-\mathbf{v}_{t}^{k}) & \text{if }t+1\notin\mathcal{I}_{E}\\
\sum_{k=1}^{N}p_{k}\left[\mathbf{v}_{t+1}^{k}+\beta_{t}(\mathbf{v}_{t+1}^{k}-\mathbf{v}_{t}^{k})\right] & \text{if }t+1\in\mathcal{I}_{E}
\end{cases}
\end{align*}
where $g_{t,k}:=\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})$ is
the stochastic gradient sampled on the $k$-th device at time $t$.  
\begin{theorem}
	\label{thm:nesterov_scvx}Let $\overline{\mathbf{v}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{v}_{T}^{k}$
	be the average of local parameters of Nesterov accelerated FedAvg
	with learning rates $\alpha_{t}=\frac{6}{\mu}\frac{1}{t+\gamma}$
	and $\beta_{t-1}=\frac{3}{14(t+\gamma)(1-\frac{6}{t+\gamma})\max\{\mu,1\}}$,
	$\gamma=\max\{32\kappa,E\}$ and $\kappa=\frac{L}{\mu}$. Then with
	full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right)
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}\leq\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right)
	\end{align*}
\end{theorem}
%
To our knowledge, this is the first convergence result for Nesterov
accelerated FedAvg in the strongly convex and smooth setting. The
same discussion about linear speedup of FedAvg applies to the Nesterov
accelerated variant. In particular, to achieve $\mathcal{O}(1/NT)$
linear speedup, $T$ iterations of the algorithm requires only $\mathcal{O}(\sqrt{NT})$
communication rounds with full participation.

@@

We now provide linear speedup analysis of Nesterov Accelerated FedAvg
with convex and smooth objectives and show that the optimality gap
is $\mathcal{O}(1/\sqrt{NT})$ where $N$ is the number of participating
devices. This result complements the strongly convex case in the previous
part, as well as the non-convex smooth setting in \cite{huo2020faster,yu2019linear,li2018federated},
where a similar $\mathcal{O}(1/\sqrt{NT})$ rate is given in terms
of averaged gradient norm. 
\begin{theorem}
	\label{thm:Nesterov_cvx}Set learning rates $\alpha_{t}=\beta_{t}=\mathcal{O}(\sqrt{\frac{N}{T}})$
	for Accelerated FedAvg with convex and $L$-smooth objectives. With
	full device participation, 
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{v}}_{t})-F^{\ast} & =\mathcal{O}(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{NT}}+\frac{NE^{2}LG^{2}}{T})
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{v}}_{t})-F^{\ast} & =\mathcal{O}(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{KT}}+\frac{E^{2}G^{2}}{\sqrt{KT}}+\frac{KE^{2}LG^{2}}{T})
	\end{align*}
\end{theorem}
%
	It is possible to extend the analyses in this section to accelerated
	FedAvg algorithms with other momentum-based updates. However, in the
	stochastic optimization setting, none of these methods will be able
	to achieve a better rate than the original FedAvg with SGD in general.
	For this reason, we will instead turn to the overparameterized setting
	\cite{ma2017power,liu2018accelerating,canziani2016analysis} in the
	next section where we show that FedAvg enjoys exponential convergence
	and it is possible to improve its convergence rate with momentum based

### 2020-06-02 source/sec_SGD.tex
updated sgd and nesterov

\subsection{Convex Smooth Objectives}
Next we provide linear speedup analyses of FedAvg with convex and
smooth objectives and show that the optimality gap is $\mathcal{O}(1/\sqrt{NT})$
where $N$ is the number of participating devices. This result complements
the strongly convex case in the previous part, as well as the nonconvex
smooth setting in \cite{jiang2018linear,yu2019parallel,haddadpour2019convergence},
where $\mathcal{O}(1/\sqrt{NT})$ results are given in terms of averaged
gradient norm. 
\begin{theorem}
	\label{thm:SGD_cvx}With full device participation and constant learning
	rate $\alpha_{t}=\mathcal{O}(\sqrt{\frac{N}{T}})$, 
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{w}}_{t})-F(\mathbf{w}^{\ast}) & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{NT}}+\frac{NE^{2}LG^{2}}{T}\right)
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round and learning rate $\alpha_{t}=\mathcal{O}(\sqrt{\frac{K}{T}})$,
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{w}}_{t})-F(\mathbf{w}^{\ast}) & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{KT}}+\frac{E^{2}G^{2}}{\sqrt{KT}}+\frac{KE^{2}LG^{2}}{T}\right)
	\end{align*}
\end{theorem}
%
\textbf{Choice of $E$ and linear speedup. }With full participation,
as long as $E=\mathcal{O}(\sqrt{T^{1/2}/N^{3/2}})$, the convergence
rate is $\mathcal{O}(1/\sqrt{NT})$ with $\mathcal{O}(N^{3/4}T^{3/4})$
communication rounds. In the partial participation setting, $E$ must
be $O(1)$ in order to achieve linear speedup of $\mathcal{O}(1/\sqrt{KT})$.
Our result again demonstrates the difference in communication complexities
between full and partial participation, and is to our knowledge the
first result on linear speedup in the general federated learning setting
with both heterogeneous data and partial participation. \\
\textbf{Learning rate. }The learning rate now depends on the final
horizon $T$ of the convergence statement, whereas in the strongly
convex case the learning rate decays as $\mathcal{O}(1/t)$. Such
a requirement $\alpha_{t}=\mathcal{O}(\sqrt{N/T})$ is also present
in the works \cite{haddadpour2019convergence,yu2019parallel} on nonconvex
problems with $\mathcal{O}(1/\sqrt{NT})$ linear speedup convergence results.

>>>>

\subsection{Convex Smooth Objectives}
Next we provide linear speedup analyses of FedAvg with convex and
smooth objectives and show that the optimality gap is $\mathcal{O}(1/\sqrt{NT})$
where $N$ is the number of participating devices. This result complements
the strongly convex case in the previous part, as well as the nonconvex
smooth setting in \cite{jiang2018linear,yu2019parallel,haddadpour2019convergence},
where $\mathcal{O}(1/\sqrt{NT})$ results are given in terms of averaged
gradient norm. 
\begin{theorem}
	\label{thm:SGD_cvx}With full device participation and constant learning
	rate $\alpha_{t}=\mathcal{O}(\sqrt{\frac{N}{T}})$, 
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{w}}_{t})-F(\mathbf{w}^{\ast}) & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{NT}}+\frac{NE^{2}LG^{2}}{T}\right)
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round and learning rate $\alpha_{t}=\mathcal{O}(\sqrt{\frac{K}{T}})$,
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{w}}_{t})-F(\mathbf{w}^{\ast}) & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{KT}}+\frac{E^{2}G^{2}}{\sqrt{KT}}+\frac{KE^{2}LG^{2}}{T}\right)
	\end{align*}
\end{theorem}
%
\textbf{Choice of $E$ and linear speedup. }With full participation,
as long as $E=\mathcal{O}(T^{1/4}/N^{3/4})$, the convergence
rate is $\mathcal{O}(1/\sqrt{NT})$ with $\mathcal{O}(N^{3/4}T^{3/4})$
communication rounds. In the partial participation setting, $E$ must
be $O(1)$ in order to achieve linear speedup of $\mathcal{O}(1/\sqrt{KT})$.
Our result again demonstrates the difference in communication complexities
between full and partial participation, and is to our knowledge the
first result on linear speedup in the general federated learning setting
with both heterogeneous data and partial participation. \\
\textbf{Learning rate. }The learning rate now depends on the final
horizon $T$ of the convergence statement, whereas in the strongly
convex case the learning rate decays as $\mathcal{O}(1/t)$. Such
a requirement $\alpha_{t}=\mathcal{O}(\sqrt{N/T})$ is also present
in the works \cite{haddadpour2019convergence,yu2019parallel} on nonconvex
problems with $\mathcal{O}(1/\sqrt{NT})$ linear speedup convergence results.
